---
title: "Making site maps in R"
author: "Claire Attridge, Stephen Brownlee"
date: "2024-04-02"
output: html_document
---

### *An introduction to shapefiles and mapping with 'sf'*

#### Lesson 1: Plotting a clean site map

</br>

#### Step 0) Pulling archived packages

This step is optional because the packages 'maptools' and 'ggsn' and 'rgdal' have been retired
These packages are useful for plotting map features including north arrows and scalebars
The functions from these packages are being translated to others, *but* this is an ongoing and incomplete process (as of 2024).
By pulling these packages from their online archives, we can still use these features in our code at present.

```{r, eval=FALSE}


url_maptools <- "https://cran.r-project.org/src/contrib/Archive/maptools/maptools_1.1-8.tar.gz"
url_ggsn <- "https://cran.r-project.org/src/contrib/Archive/ggsn/ggsn_0.5.0.tar.gz"
url_rgeos <- "https://cran.r-project.org/src/contrib/Archive/rgeos/rgeos_0.6-4.tar.gz"
url_rgdal <- "https://cran.r-project.org/src/contrib/Archive/rgdal/rgdal_1.6-7.tar.gz"
  
# repos=NULL indicates that you're not using a repository for installation
# type=source specifies that it is a source package
# This ('ggmap') is a dependency to install for ggsn
install.packages("ggmap")
install.packages(url_maptools, type="source", repos=NULL) 
install.packages(url_ggsn, type="source", repos=NULL)  
install.packages(url_rgeos, type="source", repos=NULL)  
install.packages(url_rgdal, type="source", repos=NULL)  

```

Now, we can move onto the preparation for our maps!

</br>

#### Step 1) Loading necessary packages

```{r}

library(tidyverse)
library(sf)
library(sp)
library(ggsn)
library(rgdal)
library(MetBrewer)
library(here)

```

</br>

#### Step 2) Setting the CRS and map bounds

First, we'll create a variable for the WGS84 geographic coordinate system (GCS) (ESPG: 4326).
Our map and data must first be related to a GCS before undergoing any map projections.
WGS84 is the standard global GCS that we will use to work with latlong units in decimal degrees (DD).

Next, we'll select the appropriate CRS for our study region!
Generally, for coastal BC we use the BC/Albers projected coordinate system (PCS) (ESPG: 3005).

We can project our map and data from the GCS system to this PCS system later.
As a PCS, the BC/Albers system will be specified by eastings and northings rather than latlongs.


```{r}

geog <- sf::st_crs(4326) # WGS84
proj <- sf::st_crs(3005) # BC/Albers

```

Once these variables are set, we need to decide the extent that our map will cover.
I.e., What range of latitude and longitude should it span?

This will be important for 'zooming in' on a subset area of a larger map file.


Let's explore Barkley Sound, Bamfield, in this example!


```{r}

# Setting the specifications for latitude (ymax, ymin) and longitude (xmax, xmin) ranges
# We are specifying these coordinates in DD for a GCS
ymax <- 48.922
ymin <- 48.80
xmax <- -125.05
xmin <- -125.26

```

Next, we use these specifications to create a 'blank background' for our map.
Because we specified our map corners

```{r}

bounds <- st_multipoint(matrix(c(xmax,xmin,ymax,ymin),ncol=2)) %>% 
  st_sfc(crs=geog) %>% # This sets our background in the WGS84 system
  st_sf() %>%
  st_transform(proj) # This projects our background into the BC/Albers system

# Let's plot our bounds to see where our map will go
plot(bounds)

```

</br>

#### Step 3) Reading in a shapefile

The .zip contains the shapefile of coastal British Columbia from the Hakai Institute (Daniel Okamoto, Jenn Burt), accessible via Github (https://github.com/HakaiInstitute/hakai_guide_to_r/tree/master/data/2_Shapefile).

Making use of the "here()" package can help to create a reproducible workflow.


```{r}

# First, let's check out where our current working directory is
here::here() 

# Now, we need to create a temp file for storing the .zip location
temp <- tempfile()

# We can download the .zip file by specifying the URL location and saving this to 'temp' 
URL <- "https://github.com/claireattridge/R-Tutorials/blob/main/COAST_TEST2.zip"
download.file(URL, temp)

# Then, we unzip the file so that all of the contents are open in our folder
unzip(here("COAST_TEST2.zip"))
# IMPORTANT REMINDER: Shapefiles exist as a collection of files. Aka. you need ALL of those file extensions that go by the same prefix name (e.g., COAST_TEST2) to successful map out the vector data.


# Finally we read the shapefile using an 'sf' function
# here() will specify that the .shp file is within your current directory
land <- sf::st_read(here("COAST_TEST2.shp"))


# Let's check the coordinate system of our map
st_crs(land)
# The existing CRS is WGS84 


# Let's do a quick base R plot to see what it looks like
plot(land)

```

The reason that you see multiple versions of the map, is because there are multiple "attributes" associated with the file. Let's check this out by calling the frame itself

```{r}
# First we'll check the class
class(land)
# We can see that this is already an 'sf' object that contains a dataframe of information

# Next, we'll call the shapefile object
land
# This tells us that the .shp is a multipolygon in WGS84 CRS
# The extent of the file is also specified in decimal degrees (DD)

```

Now, lets project this map from WGS84 to our BC/Albers system

```{r}

land <- land %>%
  st_transform(proj)

# Let's check the coordinate system of our map again
st_crs(land)

# It is now in BC/Albers projection!

```

Finally, let's crop this whole map to only our area of interest in Barkley Sound.

```{r}

land_crop <- land %>%
  st_crop(st_bbox(bounds)) # This specifies the extent to crop to as our previously set bounds

# Let's plot our new cropped shapefile to see if it looks correct
plot(land_crop)
# Nice! It's now only printing the zoom in of Barkley Sound.

# Now, if we want, we can save this new, cropped shapefile of our map.
# This way we don't have to do steps 1-3 every time we want to edit our map in the future.
# We would just need to load this new cropped shapefile for future work!

# Save it somewhere within your R Studio project.
# Make sure to specify the suffix as a shapefile (".shp").
write_sf(land_crop, "./land_Hakai_crop.shp", overwrite=T)

```

</br>

#### Step 4) Loading and prepping your site data

Once your map is in correct CRS and cropped to your region of study, let's pull in our site data.

You can read this is as per usual for a .csv file.
Your dataframe should at minimum contain:
i) A site identifier
ii) Latitude values
iii) Longitude values
Beyond this, you may have continuous or categorical data associated with each site in other columns.

```{r}

# For an example, let's read in data from my subtidal project sites in Barkley Sound
# We're accessing a .csv file of site locations from my Github project

urlfile <-"https://raw.githubusercontent.com/claireattridge/kelp_comms_2022/master/MSc_data/Data_new/Final_site_list_2022.csv"
sites <-read.csv(url(urlfile))

# Let's check out the dataframe to see if everything looks correct
str(sites)

# Let's make a couple of these character columns into factors for plotting later
# We'll also remove the columns that we're not interested in right now
sites_clean <- sites %>%
  dplyr::select(-c(SiteName_RLS, Original_surveyor, Temp_logger, Acoustic, Ammonium, Survey_date)) %>%
  mutate(across(where(is.character), as.factor)) %>%
  droplevels()
sites_clean

```

Our site dataframe is looking clean and ready to go.
The next step is to convert this dataframe to a 'sf' spatial object.

```{r}

# We use a function form the 'sf' package to specify what columns the coordinates are in
# The order should be your longitude column first, then latitude
sites_sf <- sites_clean %>%
  st_as_sf(coords = c(4,3))

# Now when we look at our new sf object, you'll see that a 'geometry' column has appeared
# This column specifies the data type ('POINT') and the combined coordinates as a list in each cell
sites_sf

```

This next step is sneaky but important.
Because we've recorded our GPS coordinates in decimal degrees (DD) for each site using lat/lon values,
we have to tell R that this sf object originated from a GCS coordinate system.
We will first specify the origin CRS as a GCS (in this case, WGS84), then we will project into our local PCS (BC/Albers).

```{r}

# First we set the GCS to our variable for the WGS84 code
st_crs(sites_sf) <- geog 
# Then we transform our projection into the BC/Albers code
sites_sf <- st_transform(sites_sf, proj)

# You can always double-check to make sure your CRS transformations were successful
st_crs(sites_sf)

# Let's take a look at this data on a plot
plot(sites_sf)
# Looks funny without a background reference, but it does appear that the points are reasonably spread for what we expect to see! 
# It should start to make more sense when we overlay these onto a land form and plot by a single attribute at a time.

```

</br>

#### Step 5) Putting your map together!

We're going to build our map in 'ggplot' so that we can craft the aesthetics.

```{r}

ggplot(land_crop)+
  geom_sf(fill = "grey65") +
  theme_minimal(base_size = 16) +
  geom_sf(data = sites_sf, shape=23, fill="black", size=2)

```

Starting to look good! Let's play around with the aesthetics to make it more appealing visually.
For starters, we should add a north arrow and scale bar. These are necessary map elements.

I also prefer to remove the graticules because they make a basic site location map look too busy.

```{r}

ggplot(land_crop)+
  geom_sf(fill = "grey65") +
  theme_minimal(base_size = 14) +
  geom_sf(data = sites_sf, shape=23, fill="black", size=2) + 
# Hiding the graticule lines
    theme(panel.grid.major = element_line(colour = "transparent")) +
# Adding in a north arrow 
    ggsn::north(land_crop, 
        location="topright", 
        symbol=12, 
        scale=0.1) +
# Adding in a scale bar
    ggsn::scalebar(land_crop,
               location = "bottomright",
               transform = F,
               st.bottom = F,
               st.size = 3,
               height = 0.01,
               dist = 1.5,
               dist_unit = "km",
               model = 'NAD83')

```

Great! Now we can really fiddle around to tweak the details of the axes and text.
Let's remove the unnecessary axes titles and add site labelers to each point. 

```{r}

ggplot(land_crop)+
  geom_sf(fill = "grey65") +
  theme_minimal(base_size = 14) +
  geom_sf(data = sites_sf, shape=23, fill="black", size=2) + 
# Labels
  geom_sf_text(mapping=aes(), data = sites_sf, label = sites_sf$SiteNum, stat="sf_coordinates", inherit.aes=T, size=2, nudge_y=50, nudge_x=-700) + 
# Hiding the graticule lines
    theme(panel.grid.major = element_line(colour = "transparent"),
          axis.title = element_blank(),
          axis.text = element_text(size=12, color="black"),
# We can add a black border for a clean look
          panel.border = element_rect(colour="black", fill=NA)) +
# Adding in a north arrow 
    ggsn::north(land_crop, 
        location="topright", 
        symbol=12, 
        scale=0.1) +
# Adding in a scale bar
    ggsn::scalebar(land_crop,
               location = "bottomright",
               transform = F,
               st.bottom = F,
               st.size = 3,
               height = 0.01,
               dist = 1.5,
               dist_unit = "km",
               model = 'NAD83')

```

Finally, we can color the points according to the kelp spp. composition of our forests.
This column has 4 discrete categories (Macro, Mixed, Nereo, NA) where NA indicates no forest present.

```{r}

ggplot(land_crop)+
  geom_sf(fill = "grey65") +
  theme_minimal(base_size = 14) +
# Changing our 'fill' to the column name
  geom_sf(data = sites_sf, shape=23, size=2, aes(fill=Composition)) + 
# Labels
  geom_sf_text(mapping=aes(), data = sites_sf, label = sites_sf$SiteNum, stat="sf_coordinates", inherit.aes=T, size=2, nudge_y=50, nudge_x=-700) + 
# Specifying the color palette to fill our 3 point categories
  scale_fill_manual(values=met.brewer("Egypt", 3), name="Composition") +
# Hiding the graticule lines
    theme(panel.grid.major = element_line(colour = "transparent"),
          axis.title = element_blank(),
          axis.text = element_text(size=12, color="black"),
# We can add a black border for a clean look
          panel.border = element_rect(colour="black", fill=NA)) +
# Adding in a north arrow 
    ggsn::north(land_crop, 
        location="topright", 
        symbol=12, 
        scale=0.1) +
# Adding in a scale bar
    ggsn::scalebar(land_crop,
               location = "bottomright",
               transform = F,
               st.bottom = F,
               st.size = 3,
               height = 0.01,
               dist = 1.5,
               dist_unit = "km",
               model = 'NAD83')

```

That's all for now! Stay tuned for part II where we will explore more variations to our map aesthetics, and the use of rasters and spatial calculations. 

